---
title: "Interpretation for ALL dataset SRP056295 (Pabst, Blood, 2016)"
author: "Mustafa AbuElqumsan and Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: true
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
  bookdown::html_document2: default 
# bibliography: Bibliography.bib
biblio-style: "apalike"
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, fig.path = "figures/")

```

We look forward to make the quantitative summary for some critical values and parameters, which are summaried as in the follwing table 

```{r quantitative summary}
# qunt.sammay <- data.frame( "recountID" = parameters$recountID,
#                            "Descrip.dataset" = parameters$classColumn,
#                            "no.runs" = length(loaded$runPheno$run),
#                            "No.samples before filtering" = length(loaded$runPheno$sample),
#                            "No.genes before filtering" = nrow(loaded$countsPerRuns),
#                            "No.genes NA vlues" = loaded$filteredData$naGenes,
#                            "No.genes zero.variance" = length(loaded$filteredData$zeroVarGenes),
#                            "No.genes kept filtering" = length(loaded$filteredData$keptGenes),
#                            "selected classes" = data.frame(loaded$filteredData$selectedClasses),
#                            "No.classes" = length(unique(loaded$filteredData$classes)),
#                             "Nosamp per.class"= table(loaded$samples.per.class)
#                            )

#kable(qunt.sammay)
```

A summarized report about preprocessing process before we start to study the impact of classification methods on RNAseq count data.



| recountID | No.runs | No.features | No.samples after prerocessing | No.genes after preprocessing | unique samples | pheno Table dimensions | No. of classes |
|---------------------------|--------|--------|--------|--------|--------|--------|--------|
| "SRP042620"| 167 |58037  | 162 | 55506 | 167 | 162 X 22 | 5 |
| "SRP003611"| 52  |58037 | 0 | 51239 | 8| 0 X 17 |
| "SRP061240"| 384  |58037 | 186 | 7056 | 192 | 186 X 19 | 3 |
| "SRP062966"| 117 |  58037 | 117 | 45247 | 117 | 117 X 25 | 2 |
| "SRP066834"| 729 |  58037 |  729| 15741 |729 | 729 X 23 | 3 |
| "SRP056295"| 520 | 58037 | 259  | 43780 | 263 | 259 X 17 |  2  |
